A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-16 tex/body.tex
more

@@

\begin{table*}[tbh!]
\normalsize
\begin{center}\begin{tabular}{l@{\kern1.5em}c@{\kern.5em}c@{\kern.5em}c@{\kern1.5em}c}
  & \HD{Distro} & \HD{AFP} & \HD{\gandl} & All
\MIDRULE
No (co)datatypes
  & 229 & 765 & \phantom{0}61 & 1055 \\[\jot]
%Datatype simplication only
%  &  \\
Datatypes without acyclicity
  & 235 & 770 & \phantom{0}61 & 1066 \\
Datatypes with acyclicity
  & 235 & 773 & \phantom{0}61 & 1069 \\[\jot]
%Codatatype simplication only
%  &  \\
Codatatypes without uniqueness
  & 230 & 794 & \phantom{0}65 & 1089 \\
Codatatypes with uniqueness
  & 231 & 794 & \bfseries \phantom{0}68 & 1093 \\[\jot]
Full (co)datatypes
  & \bfseries 237 & \bfseries 802 & \bfseries \phantom{0}68 & \bfseries 1107 \\[\jot]
Total number of goals
  & 879 & 2974\phantom{0} & 317 & 4170
\end{tabular}\end{center}
\caption{\,Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table*}

To evaluate the decision procedure, we generated problems from existing
Isabelle formalizations using Sledgehammer \cite{paulson-blanchette-2010}.
Codatatypes being a recent addition to Isabelle
\cite{blanchette-et-al-2014-impl}, benchmarks are somewhat hard to come by. We
included the theory files from the Isabelle distribution (Distro) and the \emph{Archive
of Formal Proofs} (AFP) \cite{klein-et-al-afp} that define codatatypes falling
within the supported fragment. We also included two unpublished theories by
Peter Gammie and Andreas Lochbihler (\gandl), about Bird and Stern--Brocot trees.
To also exercise the the support
for datatypes, the benchmarks are complemented by theories about various list
and tree data structures. The theory files were selected before running the
experiments. The experimental data is publicly available \cite{our-eval-data}.

@@

\newcommand\BAD[1]{\textcolor{red}{\textbf{#1}}}

@@

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all theories, and an overall difference of
\BAD{4.7\%}. Moreover, every aspect of the procedure, including the more
expensive rules, make a contribution. \textbf{NOT TRUE}: What is not visible in the table, but
readily available when looking at the raw data \cite{our-eval-data}, is that
the stronger decision procedures subsume the weaker ones. In the context of
Sledgehammer, the power of interpreted (co)datatypes is roughly
comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}.

%* and finally, we look in more detail at one or two such proofs, by presenting
%  it and explaining it (and, before that, understanding it)

Among the four uniqueness proofs, three corresponded to one-line
proofs in Isabelle, of the form \keyw{by}~\textit{coinduction}~\textit{auto}
\cite{blanchette-et-al-2014-impl}. The fourth proof was somewhat more elaborate:
%
\begin{quote}
\keyw{lemma} \,\textit{X0\_unique}: \,$x = \const{Node}\;0\;\const{num}\; x \Longrightarrow x = \const{X0}$ \\
\keyw{proof} \,(\textit{coinduction arbitrary}: $x$ \textit{rule}: \textit{tree.coinduct\_strong}) \\
\noindent\hbox{}\quad  \keyw{case} (\textit{Eq\_tree} $x$) \keyw{show} \textit{?case} \\
\noindent\hbox{}\qquad  \keyw{by} (\textit{subst} (1 2 3 4) \textit{Eq\_tree}) (\textit{simp add}: \textit{eqTrueI}[OF \textit{Eq\_tree}]) \\
\keyw{qed}
\end{quote}
%
\noindent
where \const{X0} is defined as $\const{X0} =
\const{Node}\;0\;\const{num}\;\const{X0}$.

@@

This work is part of a larger program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, we want to implement proof
reconstruction for CVC4's (co)datatype inferences in Isabelle. We also want to
try out CVC4 for higher-order model finding in Isabelle, as an alternative to
the counterexample generator Nitpick \cite{blanchette-nipkow-2010}. We
see some opportunities to enrich SMT solvers with recursive and corecursive
functions. Finally, it might be possible to go further in terms of supporting
nested and mixed (co)recursion and quantified formulas over (co)datatypes in
solvers.

>>>>

\newcommand\NOK[1]{\textcolor{red}{\textbf{#1}}}
\newcommand\OK[1]{#1}

@@

\begin{table*}[tbh!]
\normalsize
\begin{center}\begin{tabular}{l@{\kern1.5em}c@{\kern.5em}c@{\kern.5em}c@{\kern1.5em}c}
  & \HD{Distro} & \HD{AFP} & \HD{\gandl} & All
\MIDRULE
No (co)datatypes
  & 229 & 765 & \phantom{0}61 & 1055 \\[\jot]
%Datatype simplication only
%  &  \\
Datatypes without \rn{Acyclic}
  & 235 & 770 & \phantom{0}61 & 1066 \\
Full datatypes
  & 235 & 773 & \phantom{0}61 & 1069 \\[\jot]
%Codatatype simplication only
%  &  \\
Codatatypes without \rn{Unique}
  & 230 & 794 & \phantom{0}65 & 1089 \\
Full codatatypes
  & 231 & 794 & \bfseries \phantom{0}68 & 1093 \\[\jot]
Full (co)datatypes
  & \bfseries 237 & \bfseries 802 & \bfseries \phantom{0}68 & \bfseries 1107 \\[\jot]
Total number of goals
  & 879 & 2974\phantom{0} & 317 & 4170
\end{tabular}\end{center}
\caption{\,Number of solved goals for the three benchmark suites}
\label{tab:bench}
\end{table*}

To evaluate the decision procedure, we generated problems from existing
Isabelle formalizations using Sledgehammer \cite{paulson-blanchette-2010}.
Codatatypes being a recent addition to Isabelle
\cite{blanchette-et-al-2014-impl}, benchmarks are not abundant. We
included all the theory files from the Isabelle distribution (Distro) and the \emph{Archive
of Formal Proofs} (AFP) \cite{klein-et-al-afp} that define codatatypes falling
within the supported fragment. We also included two unpublished theories by
Peter Gammie and Andreas Lochbihler (\gandl), about Bird and Stern--Brocot trees.
To also exercise the the support
for datatypes, the benchmarks are complemented by theories about lists
and tree data structures. The theories were selected before running the
experiments. The experimental data is publicly available.%
\footnote{\url{http://TODO/FIXME.}}

@@

@@

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all theories. Overall, it accounts for a success rate
increase of \OK{4.9\%}. Moreover, every aspect of the procedure, including the
more expensive rules, make a contribution.
%* and finally, we look in more detail at one or two such proofs, by presenting
%  it and explaining it (and, before that, understanding it)
%
Three proofs were found thanks to
the \rn{Acyclic} rule and four required the \rn{Unique} rule. Among the latter,
three correspond to one-line proofs in Isabelle, of the form
\keyw{by}~\textit{coinduction}~\textit{auto} \cite{blanchette-et-al-2014-impl}.
The fourth proof, by Gammie and Lochbihler, is more elaborate:
%
\begin{quote}
\keyw{lemma} \,\textit{X0\_unique}: \,$x = \const{Node}\;0\;\const{num}\; x \Longrightarrow x = \const{X0}$ \\
\keyw{proof} \,(\textit{coinduction arbitrary}: $x$ \textit{rule}: \textit{tree.coinduct\_strong}) \\
\noindent\hbox{}\quad  \keyw{case} (\textit{Eq\_tree} $x$) \keyw{show} \textit{?case} \\
\noindent\hbox{}\qquad  \keyw{by} (\textit{subst} (1 2 3 4) \textit{Eq\_tree}) (\textit{simp add}: \textit{eqTrueI}[OF \textit{Eq\_tree}]) \\
\keyw{qed}
\end{quote}
%
\noindent
where \const{X0} is defined as $\const{X0} =
\const{Node}\;0\;\const{num}\;\const{X0}$.

The raw data also reveals that the
stronger decision procedures almost completely subsume the weaker ones. For
Sledgehammer, the power of interpreted (co)datatypes is roughly comparable to
that of arithmetic \cite{blanchette-et-al-2013-smt}.

@@

This work is part of a larger program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, it would be useful to implement
proof reconstruction for CVC4's (co)datatype inferences in Isabelle. We also
want to try out CVC4 for higher-order model finding in Isabelle, as an
alternative to the counterexample generator Nitpick
\cite{blanchette-nipkow-2010}. Finally, it might be worthwhile to enrich SMT
solvers with explicit support for recursive and corecursive functions.

%Finally, it might be possible to go
%further in terms of supporting nested and mixed (co)recursion and quantified
%ormulas over (co)datatypes in solvers.
